Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-13 paper/paper.tex
polish the appendix

\section{Hard problems and tensor networks}\label{sec:otherproblems}
\subsection{Maximal independent sets and maximal cliques}\label{sec:maximal}
Since finding maximal cliques of a graph is equivalent to finding the maximal independent sets of its complement graph, in the following,
we limit our discussion to finding maximal independent sets.
Let us denote the neighborhood of a vertex $v$ as $N(v)$ and $N[v] = N(v)\cup \{v\}$.
A maximal independent set $I_m$ is an independent set where there does exist a vertex $v \in V$ such that $I_m \cap N[v]  = \emptyset$,
i.e. an independent set that can not become a larger one by adding a new vertex.
To characterize the maximal independence restriction, one can quantify over a local set $N[v]$ with a local tensor:
\begin{equation}\label{eq:maximal}
    T(x_v, w_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx_v^{w_v} & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& \text{otherwise}.\\
    \end{cases}
\end{equation}
Intuitively, it means if all the neighbourhood vertices are not in $I_{m}$, i.e., $ s_1=s_2=\ldots=s_{|N(v)|}=0$, then $v$ should be in $I_{m}$ and contribute a factor $x_{v}$,
otherwise, if any of the neighbourhood vertices is in $I_{m}$, then $v$ cannot be in $I_{m}$.
As an example, for a vertex of degree 2, the resulting rank-3 tensor is
\begin{equation}
    T(x_v, w_v)=\left(\begin{matrix}
    \left(\begin{matrix}
        ~~0 &~1 \\
        ~~1 &~1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x_v^{w_v} &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{equation}

@@

\subsection{Max-Cut problem}
In graph theory, a cut is a partition of the vertices of a graph into two disjoint subsets,
which is also known as the boolean spin glass problem in statistic physics.
For a vertex $v\in V$, we define a boolean degree of freedom $s_v\in\{0, 1\}$.
Then the Max-Cut problem can be encoded to tensor networks by mapping an edge $(i,j)\in E$ to an edge matrix labelled by $s_is_j$
\begin{equation}
    B(x, w_{\langle i, j\rangle}) = \left(\begin{matrix}
        1 & x_{j}^{w_{\langle i,j\rangle}}\\
        x_{i}^{w_{\langle i,j\rangle}} & 1
    \end{matrix}\right),
\end{equation}
where variables $x_i$ and $x_j$ represents a cut on edge $(i, j)$ or a domain wall of an Ising spin glass.
Let $x_i^{w_{\langle i,j\rangle}} = x_j^{w_{\langle i,j\rangle}} = x$, we have a graph polynomial similar to previous ones,
where its $k$th coefficient is two times the number of configurations of cut size $k$.

\subsection{Dominating Set}
In graph theory, a dominating set for a graph $G = (V, E)$ is a subset $D \subseteq V$ such that every vertex not in $D$ is adjacent to at least one member of $D$.
The tensor network representation map a vertex $v\in V$ to a boolean degree of freedom $s_v\in\{0, 1\}$.
We defined the restriction on a vertex and its neighbouring vertices $N(v)$:
\begin{equation}
T(x, w_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
    0_v & s_1=s_2=\ldots=s_{|N(v)|}=s_v=0,\\
    1_v & s_v=0,\\
    x^{w_v} & \text{otherwise},
\end{cases}
\end{equation}
where $w_v$ is the weight of vertex $v$.
This tensor means if both $v$ and its neighbouring vertices are not in $D$, i.e., $s_1=s_2=\ldots=s_{|N(v)|}=s_v=0$,
this configuration is forbidden because $v$ is not adjacent to any member in the set.
otherwise, if $v$ is in $D$, it has a contribution $x_v^{w_v}$ to the final result.
The graph polynomial for the dominating set problem is known as the domination polynomial~\cite{Alikhani2009},
which is defined as
\begin{equation}
D(G, x) = \sum_{k=0}^{\gamma(G)} d_k x^k,
\end{equation}
where $d_k$ is the number of dominating sets of size $k$ in graph $G=(V, E)$.

\subsection{Boolean satisfiability Problem}
The boolean satisfiability problem is the problem of determining if there exists an interpretation that satisfies a given boolean formula.
One can specify a satisfiable problem in the conjuctive normal form (CNF): a conjuction of disjunctions of boolean literals,
where a disjunction of boolean literals is also called a clause.
To reduce a CNF solving to a tensor network contraction, we first map a boolean literal $a$ and its negation $\neg a$ to a same degree of freedom (label) $s_a \in \{0, 1\}$,
where 0 stands for variable $a$ having value \texttt{false} while 1 stands for having value \texttt{true}.
Then we map a clause to a tensor. For example, the $k$-th clause $\neg a \vee b \vee \neg c$ can be mapped to a tensor labeled by $(s_a, s_b, s_c)$.
\begin{equation}
C_{k} = \left(\begin{matrix}
\left(\begin{matrix}
x^{w_k} & x_{b}^{w_k} \\
x_a^{w_k} & x_{ab}^{w_k}
\end{matrix}\right) \\
\left(\begin{matrix}
x_{c}^{w_k} & x_{bc}^{w_k} \\
1 & x_{abc}^{w_k}
\end{matrix}\right)
\end{matrix}\right),
\end{equation}
where a weight is associated with a clause. There is only one entry $(s_x, s_y, s_z) = (1, 0, 1)$ that makes this clause unsatisfied.
If we contract this tensor network, we will get a multiplicative factor $x$ whenever there is a clause satisfied.
Let $x_{ab\ldots} = x$, then we can get a polynomial that its $k$-th coefficients gives us the number of assignments with $k$ clauses satisfied.

\subsection{Set packing}
Set packing is the hypergraph generalization of the maximum independent set problem, where a set corresponds to a vertex and an element corresponds to a hyperedge.
To solve the set packing problem, we can just remove the rank-2 restriction of the edge tensor in \Eq{eq:edgetensor}
\begin{equation}
    B_{v,w,\ldots, z} = \begin{cases}
        1, & v+w+\ldots+z\leq 1,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}

>>>>

\section{Hard problems and tensor networks}\label{sec:otherproblems}
\subsection{Maximal independent sets and maximal cliques}\label{sec:maximal}
Since finding maximal cliques of a graph is equivalent to finding the maximal independent sets of its complement graph, in the following,
we limit our discussion to finding maximal independent sets.
Let us denote the neighborhood of a vertex $v$ as $N(v)$ and the closed neighborhood of $v$ as $N[v] = N(v)\cup \{v\}$.
A maximal independent set $I_m$ is an independent set where there does exist a vertex $v \in V$ such that $I_m \cap N[v]  = \emptyset$,
i.e. an independent set that can not become a larger one by adding a new vertex.
To characterize the maximal independence restriction, one can quantify over a local set $N[v]$ with a local tensor:
\begin{equation}\label{eq:maximal}
    T(x_v, w_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx_v^{w_v} & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& \text{otherwise}.\\
    \end{cases}
\end{equation}
Intuitively, it means if all the neighbourhood vertices are not in $I_{m}$, i.e., $ s_1=s_2=\ldots=s_{|N(v)|}=0$, then $v$ should be in $I_{m}$ and contribute a factor $x_{v}$,
otherwise, if any of the neighbourhood vertices is in $I_{m}$, then $v$ cannot be in $I_{m}$.
As an example, for a vertex of degree 2, the resulting rank-3 tensor is
\begin{equation}
    T(x_v, w_v)=\left(\begin{matrix}
    \left(\begin{matrix}
        ~~0 &~1 \\
        ~~1 &~1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x_v^{w_v} &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{equation}

@@

\subsection{Cutting problem}
In graph theory, a cut is a partition of the vertices of a graph into two disjoint subsets,
which is also known as the boolean spin glass problem in statistic physics.
Let $G=(V,E)$ be a graph, to reduce the cutting problem on $G$ to the contraction of a tensor network, we first define a boolean degree of freedom $s_v\in\{0, 1\}$ for each vertex $v\in V$.
Then for each edge $(u,v)\in E$, we define an edge matrix labelled by $s_us_v$ as
\begin{equation}
    B(x, w_{uv}) = \left(\begin{matrix}
        1 & x_{v}^{w_{uv}}\\
        x_{u}^{w_{uv}} & 1
    \end{matrix}\right),
\end{equation}
where variables $x_u$ and $x_v$ represents a cut on this edge or a domain wall in an spin glass.
Let $x_u^{w_{uv}} = x_v^{w_{uv}} = x$, we have a graph polynomial similar to previous ones,
where its $k$th coefficient is two times the number of cuts with size $k$ (i.e. cutting $k$ edges).

\subsection{Dominating Set}
In graph theory, a dominating set for a graph $G = (V, E)$ is a subset $D \subseteq V$ such that every vertex not in $D$ is adjacent to at least one member of $D$.
To reduce this problem to the contraction of a tensor network, we first map a vertex $v\in V$ to a boolean degree of freedom $s_v\in\{0, 1\}$.
Then for each vertex $v\in V$ we define a tensor on its closed neighborhood $\{v\} \cup N(v)$ as
\begin{equation}
T(x, w_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
    0_v & s_1=s_2=\ldots=s_{|N(v)|}=s_v=0,\\
    1_v & s_v=0,\\
    x^{w_v} & \text{otherwise},
\end{cases}
\end{equation}
where $w_v$ is the weight of vertex $v$.
This tensor means if both $v$ and its neighbouring vertices are not in $D$, i.e., $s_1=s_2=\ldots=s_{|N(v)|}=s_v=0$,
this configuration is forbidden because $v$ is not adjacent to any member in the set.
Otherwise, if $v$ is in $D$, this tensor contributes a multiplicative factor $x_v^{w_v}$ to the output.
The graph polynomial for the dominating set problem is known as the domination polynomial~\cite{Alikhani2009},
which is defined as
\begin{equation}
D(G, x) = \sum_{k=0}^{\gamma(G)} d_k x^k,
\end{equation}
where $d_k$ is the number of dominating sets of size $k$.

\subsection{Boolean satisfiability Problem}
The boolean satisfiability problem is the problem of determining if there exists an interpretation that satisfies a given boolean formula.
One can specify a satisfiable problem in the conjuctive normal form (CNF): a conjuction of disjunctions of boolean literals,
where a disjunction of boolean literals is also named a clause.
To reduce the problem of solving CNF to a tensor network contraction, we first map a boolean literal $a$ and its negation $\neg a$ to a same degree of freedom (label) $s_a \in \{0, 1\}$,
where 0 stands for variable $a$ having value \texttt{false} while 1 stands for having value \texttt{true}.
Then we map a clause to a tensor. For example, the $k$-th clause $\neg a \vee b \vee \neg c$ can be mapped to a tensor labeled by $(s_a, s_b, s_c)$.
\begin{equation}
C_{k} = \left(\begin{matrix}
\left(\begin{matrix}
x^{w_k} & x_{b}^{w_k} \\
x_a^{w_k} & x_{ab}^{w_k}
\end{matrix}\right) \\
\left(\begin{matrix}
x_{c}^{w_k} & x_{bc}^{w_k} \\
1_{ac} & x_{abc}^{w_k}
\end{matrix}\right)
\end{matrix}\right),
\end{equation}
where a weight $w_k$ is associated with a clause. There is only one entry $(s_a, s_b, s_c) = (1, 0, 1)$ that makes this clause unsatisfied.
If we contract this tensor network, we will get a multiplicative factor $x$ whenever there is a clause satisfied.
Let $x^{w_k}_{\ldots} = x$, one can get a polynomial, in which the $k$-th coefficient gives the number of assignments with $k$ clauses satisfied.

\subsection{Set packing}
Set packing is the hypergraph generalization of the maximum independent set problem, where a set corresponds to a vertex and an element corresponds to a hyperedge.
To solve the set packing problem, we can just remove the rank-2 restriction of the edge tensor in \Eq{eq:edgetensor}
\begin{equation}
    B_{s_u,s_v,\ldots, s_w} = \begin{cases}
        1, & s_u+s_v+\ldots+s_w\leq 1,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}
